---
title: "Statistical methods in public policy research"
format: 
  gfm: 
    wrap: preserve
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.retina = 3
)

library(emoji)

yaml_metadata <- rmarkdown::yaml_front_matter("index.qmd")
title <- yaml_metadata$title
abstract <- yaml_metadata$abstract
```

[Andrew Heiss](https://www.andrewheiss.com) • Georgia State University

---

<!-- badges: start -->
[![HTML preprint](https://img.shields.io/badge/HTML%20preprint-FF851B)](https://stats.andrewheiss.com/snoopy-spring/) [![PDF preprint](https://img.shields.io/badge/PDF%20preprint-3D9970)](https://stats.andrewheiss.com/snoopy-spring/heiss-statistics-public-policy.pdf) [![SocArXiv Preprint](https://img.shields.io/badge/SocArXiv%20preprint-10.31235%2Fosf.io%2Fcwymb__v2-blue)](https://doi.org/10.31235/osf.io/cwymb_v2)
<!-- badges: end -->

---

> Andrew Heiss, “`{r} title`”

---

## Article summary

`{r} I(abstract)`


## `r emoji_glue(":detective::cherry_blossom:")`: Note on "snoopy spring" project name

Because project titles change all the time with revisions, rewriting, and peer review, I used [{codename}](http://svmiller.com/codename/) to generate a [NICKA-style](https://www.designation-systems.net/usmilav/codenames.html) internal-to-me project name that won't change.

```{r generate-code-name, eval=FALSE}
library(codename)
codename_message()
#> code name generated by {codename} v.0.5.0. R version 4.3.2 (2023-10-31).

codename(type = "nicka", seed = "pubpol stats")
#> [1] "snoopy spring"
```

## Licenses

**Text and figures:** All prose and images are licensed under Creative Commons ([CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)).

**Code:** All code is licensed under the [MIT License](LICENSE.md).

## Contributions and Code of Conduct

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
